Beyond Subjectivity: New Visual Tools to Choose the Best Nonlinear Dimension Reduction Layout

### 2025-06-26 paper.qmd
tinkering with title

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data (\pD{}) by applying a non-linear transformation. NLDR often exaggerates random patterns. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of \pD{} distributions. The NLDR methods and hyper-parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the \pD{} data, we have developed an algorithm to show the \gD{} NLDR model in the \pD{} space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

bibliography: paper.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand\fD{$4\text{-}D$}
---

@@

### Low density results in a squishing of the \gD{} representation in all methods {#sec-effect-dens}

Differences in density can arise by sampling at different rates in different subspaces of \pD{}. For example, the data shown in @fig-one-dens_clust-error all lies on a \gD{} curved sheet in \fD{}, but one end of the sheet is sampled densely and the other very sparsely. It was simulated to illustrate the effect of the density difference on layout generated by an NLDR, illustrated using the tSNE results.

>>>>

---
title: "Beyond Subjectivity: New Visual Tools to Choose the Best Nonlinear Dimension Reduction Layout"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data (\pD{}) by applying a non-linear transformation. NLDR often exaggerates random patterns. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of \pD{} distributions. The NLDR methods and hyper-parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the \pD{} data, we have developed an algorithm to show the \gD{} NLDR model in the \pD{} space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

bibliography: paper.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand\fD{$4\text{-}D$}
---
<!-- Looking at Non-Linear Dimension Reductions as Models in the Data Space -->

@@

### Sparseness creates a contracted \gD{} layout {#sec-effect-dens}

Differences in density can arise by sampling at different rates in different subspaces of \pD{}. For example, the data shown in @fig-one-dens_clust-error all lies on a \gD{} curved sheet in \fD{}, but one end of the sheet is sampled densely and the other very sparsely. It was simulated to illustrate the effect of the density difference on layout generated by an NLDR, illustrated using the tSNE results, but it happens with all methods.
